---
title: "Email_to_class"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MailR

```{r}
## Never worked well
## Turn on the less secure option at https://myaccount.google.com/lesssecureapps
```

## GMAILR

See 

### Testing (it works!)
```{r}
suppressPackageStartupMessages(library(gmailr))
getwd()
## edit line below to reflect YOUR json credential filename
gmailr::use_secret_file("client_secret.json")
## It would not identify itself authomatically!  Need this!!!
gmail_auth()

## edit below with email addresses from your life
test_email <- mime(
  To = "dennis.lavrov@gmail.com",
  From = "dennis.lavrov@gmail.com",
  Subject = "this is just a gmailr test",
  body = "Can you hear me now?")

send_message(test_email)

## verify that the email arrives succesfully!
```

### Creating email list

```{r}
library(tidyverse)
class_list <- read_csv("~/Downloads/SPRING2019EEOB_563.CSV") %>% select(`First Name`,`ISU EMAIL`)
#View(class_list)
class_list[3] <- sample(x=class_list[[2]],size=nrow(class_list), replace = FALSE)
class_list[4] <- sample(x=class_list[[2]],size=nrow(class_list), replace = FALSE)
same_name = TRUE
while (same_name) {
  for (i in 1:nrow(class_list)) {
    if (class_list[i,2] == class_list[i,3] || class_list[i,2] == class_list[i,4] || class_list[i,3] == class_list[i,4]) {
      class_list[3] <- sample(x=class_list[[2]],size=nrow(class_list), replace = FALSE)
      class_list[4] <- sample(x=class_list[[2]],size=nrow(class_list), replace = FALSE)
      same_name <- TRUE
      break()
    }
    else {
      same_name <- FALSE
    }
  }
}
View(class_list)

```

### Sending emails

By default gmailr will use a global project. However if you are going to be doing a heavy user and will do a lot of queries please setup your own project with the steps below.

Check https://github.com/jennybc/send-email-with-r for details!

I have found this works better -- or only works at all -- if I use Google Chrome versus, say, Safari. YMMV.

Pick a name for your project. I chose "gmailr-tutorial" for this write-up. Let's call this PROJ-NAME from now on.

Create a new project at https://console.developers.google.com/project.

Overview screen > Google Apps APIs > Gmail API.

Enable!
Click "Go to Credentials" or navigate directly to Credentials.

You want a "client ID".

Yes, you will have to "Configure consent screen".

The email should be pre-filled, since presumably you are signed in with the correct Google account. Enter PROJ-NAME as the Product name and leave everything else blank. Click Save.
Back to ... Create client ID. Select application type "Other". Enter PROJ-NAME again here for the name. Click "Create".

OAuth client pop up will show you your client ID and secret. You don't need to copy them -- there are better ways to get this info. Dismiss with "OK".

Click on the download icon at the far right of your project's OAuth 2.0 client ID listing to get a JSON file with your ID and secret. You'll get a file named something like this:

    client_secret_<LOTS OF DIGITS AND LETTERS>.apps.googleusercontent.com.json
I rename this after the project, e.g., PROJ-NAME.json and move into the directory where the bulk emailing project lives.

Optional if you are using Git, add a line like this to your .gitignore file

    PROJ-NAME.json
You can add members to specific projects from Google Developers Console, allowing them to also download JSON credentials for the same project. I do that for my TAs, for example.



```{r}
suppressPackageStartupMessages(library(gmailr))
getwd()
## edit line below to reflect YOUR json credential filename
use_secret_file("client_secret.json")
for (i in 1:nrow(class_list)) {
  recepient <- as.character(class_list[i,2])
  
  ###### MESSAGE STARTS
  text <- paste0("Dear ", class_list[i,1], ",

Please send a link to your final project GitHub repository to ",class_list[i,3]," and ", class_list[i,4]," for a peer review by 11:59pm on 4/18!  Please  remember that it is your responsibility to submit your ms. and return your reviews on time. 

Your friendly R script
:)",collapse = NULL)
  ####### MESSAGE ENDS
  
  my_email <- mime(
  To = recepient,
  Bcc = "dennis.lavrov@gmail.com",
  From = "dennis.lavrov@gmail.com",
  Subject = "Submission of the final project ms.",
  body = text)
  
  send_message(my_email)
}
```

