---
title: "Data Subsetting"
teaching: 30
exercises: 10
questions:
- "How can I work with subsets of data in R?"
objectives:
- "To be able to subset vectors, factors, lists, and dataframes"
- "To be able to extract individual and multiple elements: by index, by name, using comparison operations"
- "To be able to skip and remove elements from various data structures."
keypoints:
- "Access individual values by location using `[]`."
- "Access slices of data using `[low:high]`."
- "Access arbitrary sets of data using `[c(...)]`."
- "Use `which` to select subsets of data based on value."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
## Save the data
x <- c(5.4, 6.2, 7.1, 4.8, 7.5, 6.2)
names(x) <- c('a', 'b', 'c', 'd', 'e', 'f')
```

**Disclaimer:** This tutorial is partially based on an excellent book [Advanced R.](https://github.com/hadley/adv-r).
Read [chapter 4](https://adv-r.hadley.nz/subsetting.html) to learn how to subset other data structures.

# Subsetting  
R has many powerful subset operators and mastering them will allow you to
easily perform complex operations on any kind of dataset.

There are six different ways we can subset any kind of object, and three
different subsetting operators for the different data structures.

Let's start with the workhorse of R: atomic vectors.

## Subsetting atomic vectors

### By position

| Code      | Meaning                |
|----------:|:-----------------------|
| `x[4]`    | The fourth element     |
| `x[-4]`   | All but the forth      |
| `x[2:4]`  | Elements two to four   |
| `x[-2:4]` | All except two to four |
| `x[c(1,5)`| Elements one and five  |

> ## Vector numbering in R starts at 1
>
> In many programming languages (C and python, for example), the first
> element of a vector has an index of 0. In R, the first element is 1.
{: .callout}

### By name

| Code      | Meaning                |
|----------:|:-----------------------|
| `x["b"]`  | An element named "b"   |


> ## Tip: Duplicated names
>
> Although inexing by name is usually a much more reliable way to subset objects, 
> notice, what will happen if we have several elements named "a":
>
> ```{r}
> y <- c(5.4, 6.2, 7.1, 4.8, 7.5, 6.2)
> names(y) <- c('a', 'b', 'c', 'a', 'e', 'a')
> y[c("a", "c")]
> ```
>
> We'll solve this problem in the next section.
{: .callout}

### By position, value, or name using logical vectors

| Code                                 | Meaning                                            |
|-------------------------------------:|:---------------------------------------------------|
| `x[c(T,T,F)]`                        | Element at 1, 2, 4, 5, etc. postions               |
| `x[all.equal(x, 6.2)]`               | Element equal to 6.2                               |
| `x[x < 3]`                           | All elements less than three                       |
| `x[names(x) == "a"`                  | All elements with name "a"                         |
| `x[names(x) %in% c("a", "c", "d")`   | All elements with names "a", "c", or "d"           |
| `x[!(names(x) %in% c("a","c","d"))]` | All elements with names other than "a", "c", "d" |

> ## Discussion 1
>
>
> 1. Predict and check the results of the following operations:
> 
>    ```{r, eval=FALSE}
>    x <- c(5.4, 6.2, 7.1, 4.8, 7.5, 6.2)
>    names(x) <- c('a', 'b', 'c', 'd', 'e', 'f')
>    x[-(2:4)]
>    x[-2:4]
>    -x[2:4]
>    x[names(x) == "a"
>    ```
> 1. Discuss with your neighbor subsetting using logical values. How does each of these commands work?
>
> 1.  `x[names(x) == "a"]` `x[which(names(x) == "a")` produce the same results.
> How do these two commands differ from each other?
>
{: .discussion}

> ## Tip: Getting help for operators
>
> Remember you can search for help on operators by wrapping them in quotes:
> `help("%in%")` or `?"%in%"`.
>
{: .callout}

> ## Handling special values
>
> At some point you will encounter functions in R which cannot handle missing, infinite, or undefined data.
> 
> There are a number of special functions you can use to filter out this data:
>
> * `is.na` will return all positions in a vector, matrix, or data.frame containing `NA`.
> * likewise, `is.nan`, and `is.infinite` will do the same for `NaN` and `Inf`.
> * `is.finite` will return all positions in a vector, matrix, or data.frame that do not contain `NA`, `NaN` or `Inf`.
> * `na.omit` will filter out all missing values from a vector
{: .callout}


## Subsetting lists: three functions: `[`, `[[`, and `$`.

#### Use `[...]` to return a subset of a list as a list. 
If you want to *subset* a list, but not
*extract* an element, then you will likely use `[...]`.

```{r}
lst <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
names(lst) <- c("A","B","C","D")
lst[1]
str(lst[1])
```

#### Use `[[...]]` to *extract* an individual element of a list.

```{r}
lst[[1]]
lst[["D"]]
str(lst[[1]])
```

You **can't** extract more than one element at once **nor** use `[[ ]]` to skip elements:

```{r}
lst[[1:3]]
lst[[-1]]
```

#### Use `$...` to extract an element of a list by its name (you don't need "" for the name):

```{r}
lst$D # The `$` function is a shorthand way for extracting elements by name
```

> ## Discussion 2
>
>
> 1. What the difference between these two commands?  Why did only the second work?
> 
>    ```{r}
>    lst <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
>    names(lst) <- c("A","B","C","D")
>    # command 1:
>    lst[1][2]
>    # command 2:
>    lst[[1]][2]
>    ```
{: .discussion}

## Subsetting dataframes

Remember the data frames are lists underneath the hood, so similar rules
apply. However they are also two dimensional objects!

#### Subset columns from a dataset using `[...]`. 
Remember the data frames are lists underneath the hood, so similar rules apply.  
However, the resulting object will be a data frame:

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  z = c(TRUE,FALSE,TRUE),
  stringsAsFactors = FALSE)
str(df)
df[1]
```
#### Subset cells from a dataset using `[...]` with two <sets of> numbers:
The first argument corresponds to rows and the second to columns (either or both of them can be skipped!):

```{r}
df[1:2,2:3]
```

If we subset a single row, the result will be a data frame (because the elements are mixed types):

```{r}
df[3,]
```

But for a single column the result will be a vector (this can be changed with the third argument, `drop = FALSE`).

```{r}
df[,3]
```

#### Extract a single column using `[[...]]` or `$...` with the column name:

```{r}
df$x
#or
df[["x"]]
```

> ## Challenge 1
>
> Given the following code:
>
> ```{r}
> x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
> names(x) <- c('a', 'b', 'c', 'd', 'e')
> print(x)
> ```
>
> Come up with at least 3 different commands that will produce the following output:
>
> ```{r, echo=FALSE}
> x[2:4]
> ```
>
> After you find 3 different commands, compare notes with your neighbour. Did you have different strategies?
>
> > ## Solution to challenge 1
> >
> > ```{r}
> > x[2:4]
> > ```
> > ```{r}
> > x[-c(1,5)]
> > ```
> > ```{r}
> > x[c("b", "c", "d")]
> > ```
> > ```{r}
> > x[c(2,3,4)]
> > ```
> >
> {: .solution}
{: .challenge}

> ## Challenge 2
>
> Run the following code to define vector `x` as above:
>
> ```{r}
> x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
> names(x) <- c('a', 'b', 'c', 'd', 'e')
> print(x)
> ```
>
> Given this vector `x`, what would you expect the following to do?
>```{r, eval=FALSE}
> x[-which(names(x) == "g")]
>```
>
> Try out this command and see what you get. Did this match your expectation?
> Why did we get this result? (Tip: test out each part of the command on it's own - this is a useful debugging strategy)
>
> Which of the following are true:
>
> * A) if there are no `TRUE` values passed to `which`, an empty vector is returned
> * B) if there are no `TRUE` values passed to `which`, an error message is shown
> * C) `integer()` is an empty vector
> * D) making an empty vector negative produces an "everything" vector
> * E) `x[]` gives the same result as `x[integer()]`
>
> > ## Solution to challenge 2
> >
> > A and C are correct.
> >
> > The `which` command returns the index of every `TRUE` value in its
> > input. The `names(x) == "g"` command didn't return any `TRUE` values. Because
> > there were no `TRUE` values passed to the `which` command, it returned an
> > empty vector. Negating this vector with the minus sign didn't change its
> > meaning. Because we used this empty vector to retrieve values from `x`, it
> > produced an empty numeric vector. It was a `named numeric` empty vector
> > because the vector type of x is "named numeric" since we assigned names to the
> > values (try `str(x)` ).
> {: .solution}
{: .challenge}

> ## Challenge 3
>
> Given the following code:
>
> ```{r}
> x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
> names(x) <- c('a', 'b', 'c', 'd', 'e')
> print(x)
> ```
>
> Write a subsetting command to return the values in x that are greater than 4 and less than 7.
>
> > ## Solution to challenge 3
> >
> > ```{r}
> > x_subset <- x[x<7 & x>4]
> > print(x_subset)
> >```
> {: .solution}
{: .challenge}

