---
title: "Data Structures"
teaching: 40
exercises: 15
questions:
- "How can I read data in R?"
- "What are the basic data types in R?"
- "How do I represent categorical information in R?"
objectives:
- "To be aware of the different types of data."
- "To begin exploring data frames, and understand how it's related to vectors, factors and lists."
- "To be able to ask questions from R about the type, class, and structure of an object."
keypoints:
- "Use `read.csv` to read tabular data in R."
- "The basic data types in R are double, integer, complex, logical, and character."
- "Use factors to represent categories in R."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")

## Save the data
cats_orig <- data.frame(coat = factor(c("calico", "black", "tabby")),
                   weight = c(2.1, 5.0, 3.2),
                   likes_string = c(1, 0, 1))
cats_bad <- data.frame(coat = factor(c("calico", "black", "tabby", "tabby")),
                   weight = factor(c(2.1, 5.0, 3.2, '2.3 or 2.4')),
                   likes_string = c(1, 0, 1, 1))
cats <- cats_orig
```

**Disclaimer:** This tutorial is based on an excellent book [Advanced R.](https://github.com/hadley/adv-r)


## Data structures {#data-structures}


R's base data structures can be organised by their dimensionality (1d, 2d, or nd) and whether they're homogeneous 
(all contents must be of the same type) or heterogeneous (the contents can be of different types): 

|    | Homogeneous   | Heterogeneous |                |
|----|---------------|---------------|----------------|
| 1d | Atomic vector | List          |  <-- Vectors   |
| 2d | Matrix        | Data frame    |                |
| nd | Array         |               |                |

Almost all other objects are built upon these foundations. 
Note that R has no 0-dimensional, or scalar types. Individual numbers or strings are vectors of length one. 

Given an object, the best way to understand what data structures it's composed of is to use `str()`. `str()` is short for structure and it gives a compact, human readable description of any R data structure.

## Vectors {#vectors}

The basic data structure in R is the vector. Vectors come in two flavours: atomic vectors and lists. They have three common properties:

* Type, `typeof()`, what it is.
* Length, `length()`, how many elements it contains.
* Attributes, `attributes()`, additional arbitrary metadata.

They differ in the types of their elements: all elements of an atomic vector must be the same type, whereas the elements of a list can have different types.

### Atomic vectors

There are four common types of atomic vectors : logical, integer, double (often called numeric), and character.

Atomic vectors are usually created with `c()`, short for combine.

```{r}
dbl_var <- c(1, 2.5, 4.5)
# With the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)
# Use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)
chr_var <- c("these are", "some strings")
```

Atomic vectors are always flat, even if you nest `c()`'s: Try `c(1, c(2, c(3, 4)))`

Missing values are specified with `NA`, which is a logical vector of length 1. `NA` will always be coerced to the correct type if used inside `c()`.

#### Types and tests

Given a vector, you can determine its type with `typeof()`, or check if it's a specific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or, more generally, `is.atomic()`.

NB: `is.numeric()` is a general test for the "numberliness" of a vector and returns `TRUE` for both integer and double vectors. 

```{r}
is.numeric(int_var)
is.numeric(dbl_var)
```

#### Coercion

All elements of an atomic vector must be the same type, so when you attempt to combine different types they will be __coerced__ to the most flexible type. The coercion rules go: `logical` -> `integer` -> `double` -> `complex` -> `character`, where -> can be read as *are transformed into*. You can try to force coercion against this flow using the `as.` functions:


> ## Challenge 1
>
> Create the following vectors and predict their type:
> ```{r}
> a <- c("a", 1)
> b <- c(TRUE, 1)
> c <- c(1L, 10)
> d <- c(a, b, c)
```
>
> > ## Solution to Challenge 1
> >
> > ```{r}
> > typeof(a); typeof(b); typeof(c); typeof(d)
> > ```
> {: .solution}
{: .challenge}

When a logical vector is coerced to an integer or double, `TRUE` becomes 1 and `FALSE` becomes 0. This is very useful in conjunction with `sum()` and `mean()`

```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)
# Total number of TRUEs
sum(x)
# Proportion that are TRUE
mean(x)
```

Coercion often happens automatically. Most mathematical functions (`+`, `log`, `abs`, etc.) will coerce to a double or integer, and most logical operations (`&`, `|`, `any`, etc) will coerce to a logical. You will usually get a warning message if the coercion might lose information. If confusion is likely, explicitly coerce with `as.character()`, `as.double()`, `as.integer()`, or `as.logical()`. 

### Lists

Lists are different from atomic vectors because their elements can be of any type, including lists. You construct lists by using `list()` instead of `c()`:

```{r}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

Lists are sometimes called __recursive__ vectors, because a list can contain other lists. This makes them fundamentally different from atomic vectors.

```{r}
x <- list(list(list(list())))
str(x)
is.recursive(x)
```

`c()` will combine several lists into one. If given a combination of atomic vectors and lists, `c()` will coerce the vectors to lists before combining them. Compare the results of `list()` and `c()`:

```{r}
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)
```

The `typeof()` a list is `list`. You can test for a list with `is.list()` and coerce to a list with `as.list()`. You can turn a list into an atomic vector with `unlist()`. If the elements of a list have different types, `unlist()` uses the same coercion rules as `c()`.

Lists are used to build up many of the more complicated data structures in R. For example, both data frames and linear models objects (as produced by `lm()`) are lists:

```{r}
mtcars
is.list(mtcars)
mod <- lm(mpg ~ wt, data = mtcars)
mod
is.list(mod)
```
> ## Callout
>
> Where does these data come from?
>
> {: .callout}


> ## Discussion 1
>
>
> 1. What are the common types of atomic vector? How does a list differ from an atomic vector?
> 
> 1. What makes `is.vector()` and `is.numeric()` fundamentally different to `is.list()` and `is.character()`?
> 
> 1. Test your knowledge of vector coercion rules by predicting the output of the following uses of `c()`:
> 
>    ```{r, eval=FALSE}
>    c(1, FALSE)
>    c("a", 1)
>    c(list(1), "a")
>    c(TRUE, 1L)
>    ```
>
> 1.  Why do you need to use `unlist()` to convert a list to an 
>     atomic vector? Why doesn't `as.vector()` work? 
>
> 1. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?
> 
> 1. Why is the default missing value, `NA`, a logical vector? What's special
>    about logical vectors? (Hint: think about `c(FALSE, NA_character_)`.)
{: .discussion}

## Attributes {#attributes}

All objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of as a named list (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`.

The three most important attributes:

* Names, a character vector giving each element a name, described in 
  [names](#vector-names). 

* Dimensions, used to turn vectors into matrices and arrays, 
  described in [matrices and arrays](#matrices-and-arrays).

* Class, used to implement the S3 object system.
 
Each of these attributes has a specific accessor function to get and set values. When working with these attributes, use `names(x)`, `dim(x)`, and `class(x)`, not `attr(x, "names")`, `attr(x, "dim")`, and `attr(x, "class")`.

#### Names {#vector-names}

You can name elements in a vector in three ways:

* When creating it: 
```{r}
x <- c(a = 1, b = 2, c = 3)
x
names(x)
```

* By modifying an existing vector in place: 
```{r}
x <- 1:3; names(x) <- c("a", "b", "c")
```

* By creating a modified copy of a vector: 
```{r}
x <- 1:3
y <- setNames(x, c("a", "b", "c"))
```

Names don't have to be uniqueand not all elements of a vector need to have a name. If some names are missing, `names()` will return an empty string for those elements. If all names are missing, `names()` will return `NULL`.

```{r}
y <- c(a = 1, 2, 3)
names(y)

z <- c(1, 2, 3)
names(z)
```

You can create a new vector without names using `unname(x)`, or remove names in place with `names(x) <- NULL`.

### Factors

One important use of attributes is to define factors. A factor is a vector that can contain only predefined values, and is used to store categorical data. Factors are built on top of integer vectors using two attributes: the `class()`, "factor", which makes them behave differently from regular integer vectors, and the `levels()`, which defines the set of allowed values.

```{r}
x <- c("a", "b", "b", "a")
x
x <- factor(x)
x
class(x)
levels(x)
# You can't use values that are not in the levels
x[2] <- "c"
x
# NB: combining factors will produce unwanted results!
c(x, factor("b"))
class(c(x, factor("b")))
```

Factors are useful when you know the possible values a variable may take. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

Factors crip up all over R, and occasionally cause headaches for new R users. We'll discuss why in the next lesson.

## Matrices and arrays {#matrices-and-arrays}

Adding a `dim()` attribute to an atomic vector allows it to behave like a multi-dimensional __array__. A special case of the array is the __matrix__, which has two dimensions. Matrices are used commonly as part of the mathematical machinery of statistics. Arrays are much rarer, but worth being aware of.

Matrices and arrays are created with `matrix()` and `array()`, or by using the assignment form of `dim()`:

```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2)
# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))

# You can also modify an object in place by setting dim()
c <- 1:12
dim(c) <- c(3, 4)
c
dim(c) <- c(4, 3)
c
dim(c) <- c(2, 3, 2)
c
```

You can test if an object is a matrix or array using `is.matrix()` and `is.array()`, or by looking at the length of the `dim()`. `as.matrix()` and `as.array()` make it easy to turn an existing vector into a matrix or array.

Vectors are not the only 1-dimensional data structure. You can have matrices with a single row or single column, or arrays with a single dimension. They may print similarly, but will behave differently. The differences aren't too important, but it's useful to know they exist in case you get strange output from a function (`tapply()` is a frequent offender). As always, use `str()` to reveal the differences.

```{r}
str(1:3)                   # 1d vector
str(matrix(1:3, ncol = 1)) # column vector
str(matrix(1:3, nrow = 1)) # row vector
str(array(1:3, 3))         # "array" vector
```

While atomic vectors are most commonly turned into matrices, the dimension attribute can also be set on lists to make list-matrices or list-arrays:

```{r}
l <- list(1:3, "a", TRUE, 1.0)
dim(l) <- c(2, 2)
l
```

These are relatively esoteric data structures, but can be useful if you want to arrange objects into a grid-like structure. For example, if you're running models on a spatio-temporal grid, it might be natural to preserve the grid structure by storing the models in a 3d array.

> ## Discussion 2
>
> 1.  What does `dim()` return when applied to a vector?
> 
> 1.  If `is.matrix(x)` is `TRUE`, what will `is.array(x)` return?
> 
> 1.  How would you describe the following three objects? What makes them
>     different to `1:5`?
>
>    ```{r}
>    x1 <- array(1:5, c(1, 1, 5))
>    x2 <- array(1:5, c(1, 5, 1))
>    x3 <- array(1:5, c(5, 1, 1))
>    ```
{: .discussion}

## Data frames {#data-frames}

A data frame is the most common way of storing data in R, and if [used systematically](http://vita.had.co.nz/papers/tidy-data.pdf) makes data analysis easier. Under the hood, a data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list.  This means that a data frame has `names()`, `colnames()`, and `rownames()`, although `names()` and `colnames()` are the same thing. The `length()` of a data frame is the length of the underlying list and so is the same as `ncol()`; `nrow()` gives the number of rows.

As described in [subsetting](#subsetting), you can subset a data frame like a 1d structure (where it behaves like a list), or a 2d structure (where it behaves like a matrix).

### Creation

You create a data frame using `data.frame()`, which takes named vectors as input:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)
```

Beware `data.frame()`'s default behaviour which turns strings into factors. Use `stringAsFactors = FALSE` to suppress this behaviour! Compare:

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE)
str(df)
```

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"))
str(df)
```

### Testing and coercion

Because a `data.frame` is an S3 class, its type reflects the underlying vector used to build it: the list. To check if an object is a data frame, use `class()` or test explicitly with `is.data.frame()`:

```{r}
typeof(df)
class(df)
is.data.frame(df)
```

You can coerce an object to a data frame with `as.data.frame()`:

* A vector will create a one-column data frame.

* A list will create one column for each element; it's an error if they're 
  not all the same length.
  
* A matrix will create a data frame with the same number of columns and rows as the matrix.

> ## Discussion 3
>
> 1.  What attributes does a data frame possess?
> 
> 1.  What does `as.matrix()` do when applied to a data frame with 
>     columns of different types?
> 
> 1.  Can you have a data frame with 0 rows? What about 0 columns?
{: .discussion}

### Reading data into a dataframe

One of R's most powerful features is its ability to deal with tabular data -
like what you might already have in a spreadsheet or a CSV. 
The `read.csv` function is used for reading in tabular data stored in a text
file where the columns of data are delimited by commas (csv = comma separated
values). Tabs are also commonly used to separated columns - if your data are in
this format you can use the function `read.delim`. If the columns in your data
are delimited by a character other than commas or tabs, you can use the more
general and flexible `read.table` function. CSV data are read into dataframes.

Let's start by making a toy dataset in your `data/` directory, called `feline-data.csv` with the following data:

```{r, eval=FALSE}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

> ## Tip: Editing Text files in R
>
> You can create `data/feline-data.csv` using a text editor (vi or Nano),
> or within RStudio with the **File -> New File -> Text File** menu item.
{: .callout}

We can load this into R via the following:

```{r}
cats <- read.csv(file = "data/feline-data.csv")
cats
```

We can begin exploring our dataset right away, pulling out columns by specifying
them using the `$` operator:

```{r}
cats$weight
cats$coat
```

We can do other operations on the columns:

```{r}
## Say we discovered that the scale weighs two Kg light:
cats$weight + 2
paste("My cat is", cats$coat)
```

We can ask what type of
data something is:

```{r}
typeof(cats$weight)
```


> ## Challenge 2
>
> There are several subtly different ways to call variables, observations and
> elements from data.frames:
>
> - `cats[1]`
> - `cats[[1]]`
> - `cats$coat`
> - `cats["coat"]`
> - `cats[1, 1]`
> - `cats[, 1]`
> - `cats[1, ]`
>
> Try out these examples and explain what is returned by each one.
>
> *Hint:* Use the function `typeof()` to examine what is returned in each case.
>
> > ## Solution to Challenge 2
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1]
> > ```
> > We can think of a data frame as a list of vectors. The single brace `[1]`
> returns the first slice of the list, as another list. In this case it is the
> first column of the data frame.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[[1]]
> > ```
> > The double brace `[[1]]` returns the contents of the list item. In this case
> it is the contents of the first column, a _vector_ of type _factor_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats$coat
> > ```
> > This example uses the `$` character to address items by name. _coat_ is the
> first column of the data frame, again a _vector_ of type _factor_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats["coat"]
> > ```
> > Here we are using a single brace `["coat"]` replacing the index number with
> the column name. Like example 1, the returned object is a _list_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1, 1]
> > ```
> > This example uses a single brace, but this time we provide row and column
> coordinates. The returned object is the value in row 1, column 1. The object
> is an _integer_ but because it is part of a _vector_ of type _factor_, R
> displays the label "calico" associated with the integer value.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[, 1]
> > ```
> > Like the previous example we use single braces and provide row and column
> coordinates. The row coordinate is not specified, R interprets this missing
> value as all the elements in this _column_ _vector_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1, ]
> > ```
> > Again we use the single brace with row and column coordinates. The column
> coordinate is not specified. The return value is a _list_ containing all the
> values in the first row.
> {: .solution}
{: .challenge}


> ## Challenge 3
>  Create a list of length two containing a character vector for each of the sections in this part of the workshop:
>
>  - Data types
>  - Data structures
>
>  Populate each character vector with the names of the data types and data
>  structures we've seen so far.
>
> > ## Solution to Challenge 3
> > ```{r}
> > dataTypes <- c('double', 'complex', 'integer', 'character', 'logical')
> > dataStructures <- c('data.frame', 'vector', 'factor', 'list', 'matrix')
> > answer <- list(dataTypes, dataStructures)
> > ```
> > Note: it's nice to make a list in big writing on the board or taped to the wall
> > listing all of these types and structures - leave it up for the rest of the workshop
> > to remind people of the importance of these basics.
> >
> {: .solution}
{: .challenge}