---
title: Developing Workflows with R Scripts
teaching: 45
exercises: 20
questions:
- "How can I make data-dependent choices in R?"
- "How can I repeat operations in R?"
objectives:
- "Write conditional statements with `if()` and `else()`."
- "Write and understand `for()` loops."
keypoints:
- "Use `if` and `else` to make choices."
- "Use `for` to repeat operations."
---

```{r, include=FALSE}
if (!require("microbenchmark")) install.packages("microbenchmark")
source("../bin/chunk-options.R")
knitr_fig_path("07-")
# Silently load in the data so the rest of the lesson works
mtfs <- read.delim("https://raw.githubusercontent.com/vsbuffalo/bds-files/master/chapter-08-r/motif_recombrates.txt", header=TRUE)
# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

## Developing Workflows with R Scripts

### Control Flow: if, for, and while

> ## Before we start
> If you programmed before in other languages, you'll be happy 
> to finally see a section on control flow. However, before you 
> start writing loops in R, I encourage you to look back at some 
> of the functions we used already, particularly `split` and 
> `lapply`. These functions can replace many of the loops with 
> a cleaner and faster code. 
> {: .callout}

Often we want to control the flow of our actions. This can be done
by setting actions to occur only if a condition or a set of conditions are met.
Alternatively, we can also set an action to occur a particular number of times.

Three common control flow and looping statments are `if`, `for`, and `while`:

```{r, eval=FALSE}
# if
if (x == some_value) {
            # do some stuff in here
} else if (x == other_value) {
            # elseif is optional
} else {
            #else is optional
}

#for
for (element in some_vector) {
            # iteration happens here
}

#while
while (something_is_true) {
            # do some stuff
}
```

You can break out of for and while loops with a `break` statement, and advance loops to the next iteration with `next`. 

#### Examples:

```{r}
# sample a random number from a Poisson distribution
# with a mean (lambda) of 8

x <- rpois(1, lambda=8)

if (x >= 10) {
  print("x is greater than or equal to 10")
}
x
```
  
> ## Tip: pseudo-random numbers
>
> In the above case, the function `rpois()` generates a random number following a
> Poisson distribution with a mean (i.e. lambda) of 8. We can use function `set.seed()`
> guarantees that all machines will generate the exact same 'pseudo-random'
> number ([more about pseudo-random numbers](http://en.wikibooks.org/wiki/R_Programming/Random_Number_Generation)).
>
{: .callout}

```{r}
for(i in 1:10){
  print(i)
}
```

```{r}
for(i in 1:3){
  for(j in c('a', 'b', 'c')){
    print(paste(i,j))
  }
}
```

```{r}
z <- 1
while(z > 0.1){
  z <- runif(1)
  print(z)
}
```

#### `ifelse`

R also has a vectorized version of if: the `ifelse` function. Rather than control program flow, ifelse(test, yes, no) 
returns the yes value for all TRUE cases of test, and no for all FALSE cases. For example:

```{r}
x <-c(-3,1,-5,2)
ifelse(x < 0, -1, 1)
```

Note, that although `ifelse()` is readable and clear, it can be slower than the following:

```{r}
x <- c(-3, 1, -5, 2)
y <- rep(1, length(x))
y[x < 0] <- -1
y
```

> ## Pre-allocating vectors
>
> One point worth mentioning is that if you do need to loop over a structure
> with a `for` or `while` loop, always *preallocate* your results vector. You can
> create empty numeric vectors with `numeric(len)` and empty integer vectors with
> `integer(len)`, where `len` is the length. For example, if you were to loop
> over too vectors to sum their values pairwise, do not use:
>
> ```{r}
> x <- rnorm(10)
> y <- rnorm(10)
> res <- NULL
> for (i in 1:length(x)) {
>     res <- c(res, x[i] + y[i])
> }
> ```
> First, this is completely unecessary since we could calculate this with `res <-
> x + y`. Second, this code would be needlessly slow because appending to vectors with `res <-
> c(res, ...)` is extremely computationally expensive in R.
>
> R's vectors are allocated in memory to the _exact_ length that the need to be.
> So each iteration of the above code requires (1) allocating a new `res` vector
> one element longer, and (2) copying all of `res`'s current elements over to
> this new vector. 
>
> The correct way to do this (assuming in your real code something in the `for`
> loop can't be parallelized!) is:
>
> ```{r}
> x <- rnorm(10)
> y <- rnorm(10)
> res <- numeric(10) # preallocation!
> for (i in 1:length(x)) {
>     res[i] <- x[i] + y[i]
> }
> ```
{: .callout}

### Writing functions
R, at its heart, is a functional programming (FP) language. 
This means that it provides many tools for the creation and 
manipulation of functions. 

The most important thing to understand about functions in R  
is that they are objects in their own right. You can work 
with them exactly the same way you work with any other type 
of object.

All R functions have three parts:  
* the `body()`, the code inside the function.
* the `formals()`, the list of arguments which controls how you can call the function.
* the `environment()`, the “map” of the location of the function’s variables.

When you print a function in R, it shows you these three components. 
If the environment isn’t displayed, it means that the function was 
created in the global environment.

R makes writing functions very easy, both because you should 
be writing lots of them to organize your code and applying 
functions is such a common operation in R.

The general syntax for R functions is:  

```{r}
fun_name <- function(args) {
# body, containing R expressions 
return(value)
}
```

In some cases you can use a simplified format:

```{r}
fun_name <- function(args) #body
```

For example, 

```{r}
meanRemoveNA <- function(x) mean(x, na.rm=TRUE)
```

Functions that contain only one line in their body can omit 
the braces (as the meanRemoveNA() function does). Similarly, 
using `return()` to specify the return value is optional; 
R’s functions will automatically return the last evaluated 
expression in the body.

We could forgo creating a function named `meanRemoveNA()` in 
our global environment altogether and instead use an anonymous 
function (named so because anonymous functions are functions 
without a name). Anonymous functions are useful when we only 
need a function once for a specific task. For example, if we 
wrote meanRemoveNA() to use with lapply, we could write:

```{r}
lapply(ll, function(x) mean(x, na.rm=TRUE))
```

> ## Illustration of the importance of preallocation
>
> Now that we know how to write functions, I can illustrate 
> the importance of preallocation.  Let's start by converting 
> our previous code into functions:
> ```{r}
> sum-no-preallocation <- function(x,y) {
>   res <- NULL
>   for (i in 1:length(x)) {
>       res <- c(res, x[i] + y[i])
>   }
> }
> sum-with-preallocation <- function(x,y) {
>   res <- numeric(length(x)) # preallocation!
>   for (i in 1:length(x)) {
>       res[i] <- x[i] + y[i]
>   }
> }
> # Let's test these functions!
> x <- rnorm(100)
> y <- rnorm(100)
> library(microbenchmark)
> microbenchmark(sum-no_preallocation(x,y), sum-with-preallocation(x,y), x+y)
> ```
{: .callout}


### Working with R Scripts

Although we’ve used R interactively, in practice your analyses should be kept in scripts that can be run 
many times throughout development. Scripts can be organized into project directories and checked into Git 
repositories. There’s also a host of excellent R tools (e.g., Knitr and Rmarkdown) to help in creating 
well-documented, reproducible projects in R.

You can run R scripts from R using the function `source()`. For example, to execute an R script named _my_analysis.R_ 
use:

```{r}
source("my_analysis.R")
```

Alternatively, we can execute a script in batch mode from the command line with:

```{r, eval=FALSE}
$ Rscript --vanilla my_analysis.R
```

This comes in handy when you need to rerun an analysis on different files or with different arguments 
provided on the command line. It's a good practice to use --vanilla option because by default, Rscript 
will restore any past saved environments and save its current environment after the execution completes. 
Usually we don’t want R to restore any past state from previous runs, as this can lead to irreproducible 
results. Additionally, saved environments can make it a nightmare to debug a script. See R --help for 
more information.

Lastly, if you want to retrieve command-line arguments passed to your script, use R’s commandArgs() 
with trailingOnly=TRUE. For example, this simple R script just prints all arguments:

```{r, eval=FALSE}
## args.R -- a simple script to show command line args
args <- commandArgs(TRUE) 
print(args)
```

We run this with:

```{r, eval=FALSE}
$ Rscript --vanilla args.R arg1 arg2 arg3
```
  
> ## Working directory
> It’s important to mind R’s working directory when writing scripts. Scripts should not use setwd() to set 
> their working directory, as this is not portable to other systems (which won’t have the same directory 
> structure). For the same reason, use relative paths like data/achievers.txt when loading in data, and not 
> absolute paths like /Users/jlebowski/data/achievers.txt. Also, it’s a good idea to indicate (either in comments 
> or a README file) which directory the user should set as their working directory.
{: .callout}

### Workflows for Loading and Combining Multiple Files

See pages 257-260 of the book if you need to load and combine multiple files

### Exporting Data

At some point during an analysis, you’ll need to export data from R. We can export dataframes to plain-text files 
using the R function `write.table()`. Unfortunately, `write.table()` has some poorly chosen defaults that we usually 
need to adjust. For example, if we wanted to write our dataframe mtfs to a tab-delimited file named hotspot_motifs.txt, 
we would use:

```{r}
write.table(mtfs, file="hotspot_motifs.txt", quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
```

`write.table()`’s first two arguments are the dataframe (or matrix) to write to file and the file path to save it to. 
By default, `write.table()` quotes factor and character columns and includes rownames. Qe disable these with `quote=FALSE` and `row.names=FALSE`. We also can set the column separators to tabs with the `sep` argument. A header can be included or excluded with the `col.names` argument.

Given that many Unix tools work well with Gzipped files, it may be useful to write a compressed version of a 
file directly from R. In addition to a string file path, `write.table`’s file argument also handles open file 
connections. So to write our dataframe to a compressed file, we can open a gzipped file connect with gzfile():

```{r}
hs_gzf <- gzfile("hotspot_motifs.txt.gz")
write.table(mtfs, file=hs_gzf, quote=FALSE, sep="\t", row.names=FALSE, col.names=TRUE)
```

While plain-text formats are the preferable way to share tabular datasets, it’s not the best way to save 
complex R data structures like lists or special R objects. In these cases, it’s usually preferable to save R 
objects as R objects. Encoding and saving objects to disk in a way that allows them to be restored as the original 
object is known as serialization. R’s functions for saving and loading R objects are `save()` and `load()`. 

```{r}
tmp <- list(vec=rnorm(4), df=data.frame(a=1:3, b=3:5))
save(tmp, file="example.Rdata")
rm(tmp) # remove the original 'tmp' list
load("example.Rdata") # this fully restores the 'tmp' list from file
str(tmp)
```

The function `save()` has an analogous function called `save.image()` that saves all objects in your workspace 
rather than just the objects you specify. Combined with savehistory(), save.image() can be a quick way to store 
your past work in R in a rush.